Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-01 paper_jasa.tex
cleaned up section 4, needs some additions as indicated

Theoretical power for the regression parameters is derived in the next section.

Consider a linear regression model 
\begin{equation}\label{multi} Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i 
\end{equation}
where $\epsilon_i \stackrel{iid}{ \sim } N(0,\sigma^2)$, $i=1,2, .., n$. The covariates ($X_j, j=1,..,p$) can be continuous or discrete.

In a regression setting, there is a multitude of established graphics that help with the evaluation and diagnostics of a model \citep{cook:99}. Table \ref{tbl:stat_multiple} shows an overview of some commonly used charts and their corresponding hypothesis tests. Note that some charts can be associated with several different testing situations. In our paper, we pick the examples for situation 2 and 3 as candidates.

A lineup including the test statistic  for a binary $X_k$ is shown in Figure \ref{fig:test_category}. The 19 null plots are generated  by simulating residuals from $N(0,\hat{\sigma}^2)$. The test statistic, the plot containing the actual data, is randomly placed among these null plots. If the test statistic is identifiable the null hypothesis is rejected. % with a $p$-value of at most 0.05.

Figure \ref{fig:power_expected} shows the power of the uniformly most powerful test (UMP) in comparison to the expected power of the visual test obtained from equation \eqref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of UMP test. In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

\subsection{Estimating Power from Lineups}

@@

>>>>

Theoretical power for the regression parameters is derived in the next section.
\subsection{Estimating Power from Lineups}

@@

To make these concepts more concrete consider how this would operate in the linear models setting. Consider a linear regression model 
\begin{equation}\label{multi} Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i 
\end{equation}
where $\epsilon_i \stackrel{iid}{ \sim } N(0,\sigma^2)$, $i=1,2, .., n$. The covariates ($X_j, j=1,..,p$) can be continuous or discrete.

In this setting, there is a multitude of established graphics that help with the evaluation and diagnostics of a model \citep{cook:99}. Table \ref{tbl:stat_multiple} shows an overview of possible hypotheses that would be tested in the regression setting, and what visual test statistic might be used in these tests. These plots shoul be familiar, because they are all commonly used for evaluating and diagnosing regression models. For example, to examine the effect of variable $X_j$ on $Y$, we would plot $Y$ against $X_j$ (cases 1-4). To assess whether the linear assumption is appropriate we would make a plot of residuals against fitted values (case 5).

%Note that some charts can be associated with several different testing situations. In our paper, we pick the examples for situation 2 and 3 as candidates.

For the purposes of validating visual inference, we focus on cases 2 and 3, with a continuous and catgorical explanatory variable respectively.

%A lineup including the test statistic  for a binary $X_k$ is shown in Figure \ref{fig:test_category}. The 19 null plots are generated  by simulating residuals from $N(0,\hat{\sigma}^2)$. The test statistic, the plot containing the actual data, is randomly placed among these null plots. If the test statistic is identifiable the null hypothesis is rejected. % with a $p$-value of at most 0.05.

Figure \ref{fig:power_expected} shows the power of the uniformly most powerful test (UMP) in comparison to the expected power of the visual test obtained from equation \eqref{power_exp}. \green{Give more details on what is the UMP test in this setting.} Notice that the expected power of the visual test is almost as good as the power of UMP test. In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.
